|

### 2021-08-01 prova_knit_finale.Rmd
change
non va SVAPFopt

    \item Compute: $m^n=a\cdot  \psi^n+(1-a)\cdot \bar \psi$ for $n=1,...,N$, with $a\in (0,1)$ and $h^2=1-a^2$.

@@

If we instead want to sample from the optimal transition kernel, things get more complicated. 
It can be shown that the optimal transition kernel is
\[ m_{t}^{opt}(x_{t}|x_{t-1}) \propto f_{t}(y_{t}|x_{t})p(x_{t}|x_{t-1}) \propto exp\bigg[-\frac{1}{2\sigma^{2}}(x_{t}-\alpha-\beta x_{t-1})^{2}-\frac{e^{-x_{t}}}{2}y_{t}^{2}-\frac{x_{t}}{2}\bigg]\]
and it is not straightforward to sample from it. Therefore, we follow a common praxis consisting in approximating $m_{t}^{opt}$ by a Gaussian distribution obtained by linearising $e^{-x_{t}}$, i.e. $e^{-x_{t}}=e^{-x_{0}}(1+x_{0}-x+...)$. Taking $x_{0}=\mu^{\star}(x_{t-1}) := \alpha + \beta x_{t-1}$, i.e the mean of the Gaussian transition density $p(x_{t}|x_{t-1})$. After some calculations we would end up with a Gaussian distribution $N(\xi(x_{t-1}),\sigma^{2})$ with 
\[ \xi(x_{t-1}) = \mu^{\star}(x_{t-1}) + \frac{\sigma^{2}}{4}(y_{t}^{2}e^{-\mu^{\star}(x_{t-1})}-2)\].
With this transition kernel we can implement the Guided Particle Filter in the following way

@@

The Auxiliary Particle Filter with optimal transition kernel uses the same transition kernel of the Guided Particle Filter and we decided to use an auxiliary function $g(x_{t-1})=E(x_{t}|x_{t-1})=\alpha+\beta x_{t-1}$.

\begin{algorithm}APF with optimal transition kernel for Stochastic Volatility Model
\begin{itemize}
\item Let $\{(x_{0},w_{0})^{(i)}\}_{i=1}^{N}$ summarizes $p(x_{0}|y_{0})$ such that, for example, $E(g(x_{0})|y_{0}) \approx \sum_{i=1}^{N}w_{0}^{(i)}g(x_{0}^{(i)})$. In particular, initialize $(x_{0}^{(1)},...,x_{0}^{(N)})$ from $N(m_{0},C_{0})$ and set $w_{0}^{(i)}=N^{-1} \ \forall \ i=1,...,N$.
\item For $t=1,...,n$:
\begin{enumerate}
\item Compute $ESS=\Bigg(\sum_{i=1}^{N}(w_{t}^{(i)})^{2}\Bigg)^{-1}$
\item if $ESS<N/2$ then
\begin{enumerate}
\item Draw $I_{k}$ with $P(I_{k}=i) \propto w_{t-1}^{(i)}f_{N}(y_{t}|0,e^{\alpha+\beta x_{t-1}}) $
\item $x_{t}^{(i)} \sim N(\alpha+\beta x_{t-1}^{(I_{k})}+\frac{\tau^2}{4}\big[y_{t}^2e^{-(\alpha+\beta x_{t-1}^{(I_{k})})}-2\big],\tau^{2})$ for $i=1,...N$
\item Set  $\tilde{w}^{(i)}_{t} = N^{-1}$ for $i=1,...,N$
\end{enumerate}
\item otherwise
\begin{enumerate}
\item Draw $x_{t}^{(i)} \sim N(\alpha+\beta x_{t-1}+\frac{\tau^2}{4}\big[y_{t}^2e^{-(\alpha+\beta x_{t-1})}-2\big],\tau^{2}) \ \ i=1,...,N$ for $i=1,...N$
\item Set $\tilde{w}^{(i)}_{t}=w^{(i)}_{t-1}f_{N}(y_{t};0,\alpha+\beta x_{t})$ for $i=1,...,N$
\end{enumerate}
\item Normalize the weights: $w_{t}^{(i)}=\frac{\tilde{w}_{t}^{(i)}}{\sum_{j=1}^{N}(\tilde{w}_{t}^{(j)})}$
\end{enumerate}
\end{itemize}
\end{algorithm}

@@

@@

    xprev<-x
    x<-rnorm(N,alpha+beta*x+(tau^2)/4*((data[t]^2)*exp(-(alpha+beta*x))-2),tau)
    w1<-w*dnorm(data[t],0,exp(x/2))*dnorm(x,alpha+beta*xprev,tau)/
      dnorm(x,mean=alpha+beta*xprev+(tau^2)/4*((data[t]^2)*exp(-(alpha+beta*xprev))-2),tau)

@@

  for(t in 1:length(data)){
    x_prev=x
    x_hat = alpha+beta*x #means of the transition from the previous particle 
    I_k = sample(1:N,prob=w*dnorm(data[t],0,sqrt(exp((x_hat)/2))))
    for(k in 1:N){
      x[k]=rnorm(size=1,mean=alpha+beta*x_prev[I_k],sd=tau)
      w[k]=dnorm(data[t],0,sqrt(exp((x[k]))))/dnorm(data[t],0,sqrt(exp((x_hat))))
    }
    w = w/sum(w)
    ESS  = 1/sum(w^2)

@@

>>>>

    \item Compute: $m^{A_t^{n}}=a \cdot  \psi^{{A_t}^{n}}+(1-a)\cdot \bar \psi$ for $n=1,...,N$, with $a\in (0,1)$ and $h^2=1-a^2$.

@@

If we instead want to sample from the optimal transition kernel, things get more complicated. 
It can be shown that the optimal transition kernel is
\[ m_{t}^{opt}(x_{t}|x_{t-1}) \propto f_{t}(y_{t}|x_{t})p(x_{t}|x_{t-1}) \propto exp\bigg[-\frac{1}{2\sigma^{2}}(x_{t}-\alpha-\beta x_{t-1})^{2}-\frac{e^{-x_{t}}}{2}y_{t}^{2}-\frac{x_{t}}{2}\bigg]\]
and it is not straightforward to sample from it. Therefore, we follow a common praxis consisting in approximating $m_{t}^{opt}$ by a Gaussian distribution obtained by linearising $e^{-x_{t}}$, i.e. $e^{-x_{t}}=e^{-x_{0}}(1+x_{0}-x+...)$, taking $x_{0}=\mu^{\star}(x_{t-1}) := \alpha + \beta x_{t-1}$, i.e the mean of the Gaussian transition density $p(x_{t}|x_{t-1})$. After some calculations we would end up with a Gaussian distribution $N(\xi(x_{t-1}),\sigma^{2})$ with 
\[ \xi(x_{t-1}) = \mu^{\star}(x_{t-1}) + \frac{\sigma^{2}}{4}(y_{t}^{2}e^{-\mu^{\star}(x_{t-1})}-2)\].
With this transition kernel we can implement the Guided Particle Filter in the following way

@@

The Auxiliary Particle Filter with optimal transition kernel uses the same transition kernel of the Guided Particle Filter and while the perfectly adapted auxiliary function is
\[g(x_{t-1})=\frac{1}{2\pi\sigma^{2}}\int_{\mathcal{X}}exp\bigg[-\frac{1}{2\sigma^{2}}(x_{t}-\alpha-\beta x_{t-1})^{2}-\frac{e^{-x_{t}}}{2}y_{t}^{2}-\frac{x_{t}}{2}\bigg] dx_{t-1}\]
The relevant literature suggest to approximate this integral by linearising the exponential term as in the GPF case. The resulting auxiliary function is
\[ g(x_{t-1})=exp \bigg[ \frac{1}{2\sigma^{2}} \big\{\xi(x_{t-1})^{2}-\mu^{\star}(x_{t-1})^{2} \big\} - \frac{y_{t}^{2}}{2}e^{-\mu^{\star}(x_{t-1})} \big\{ 1+\mu^{\star}(x_{t-1}) \big\}  \bigg]\]
where $\xi(x_{t-1})$ and $\mu^{\star}$ are defined as in the case of GPF.

\begin{algorithm}APF with optimal transition kernel for Stochastic Volatility Model
\begin{itemize}
\item Let $\{(x_{0},w_{0})^{(i)}\}_{i=1}^{N}$ summarizes $p(x_{0}|y_{0})$ such that, for example, $E(g(x_{0})|y_{0}) \approx \sum_{i=1}^{N}w_{0}^{(i)}g(x_{0}^{(i)})$. In particular, initialize $(x_{0}^{(1)},...,x_{0}^{(N)})$ from $N(m_{0},C_{0})$ and set $w_{0}^{(i)}=N^{-1} \ \forall \ i=1,...,N$.
\item For $t=1,...,n$:
\begin{enumerate}
\item Compute $ESS=\Bigg(\sum_{i=1}^{N}(w_{t}^{(i)})^{2}\Bigg)^{-1}$
\item if $ESS<N/2$ then
\begin{enumerate}
\item Draw $I_{k}$ with $P(I_{k}=i) \propto w_{t-1}^{(i)}g(x_{t-1}) $
\item $x_{t}^{(i)} \sim N(\alpha+\beta x_{t-1}^{(I_{k})}+\frac{\tau^2}{4}\big[y_{t}^2e^{-(\alpha+\beta x_{t-1}^{(I_{k})})}-2\big],\tau^{2})$ for $i=1,...N$
\item Set  $\tilde{w}^{(i)}_{t} = N^{-1}$ for $i=1,...,N$
\end{enumerate}
\item otherwise
\begin{enumerate}
\item Draw $x_{t}^{(i)} \sim N(\alpha+\beta x_{t-1}+\frac{\tau^2}{4}\big[y_{t}^2e^{-(\alpha+\beta x_{t-1})}-2\big],\tau^{2}) \ \ i=1,...,N$ for $i=1,...N$
\item Set $\tilde{w}^{(i)}_{t}=w^{(i)}_{t-1}g(x_{t-1})$ for $i=1,...,N$
\end{enumerate}
\item Normalize the weights: $w_{t}^{(i)}=\frac{\tilde{w}_{t}^{(i)}}{\sum_{j=1}^{N}(\tilde{w}_{t}^{(j)})}$
\end{enumerate}
\end{itemize}
\end{algorithm}

@@

```{r message=FALSE, warning=FALSE, include=FALSE, paged.print=FALSE}
SVPFfun<-function(data,N,m0,C0,alpha,beta,tau,r){
  if(missing(r)){r=2}else{}
  xs<-NULL
  ws<-NULL
  ess<-NULL
  x  = rnorm(N,m0,sqrt(C0))
  w  = rep(1/N,N)

    xs = rbind(xs,x)
    ws = rbind(ws,w)
    ess =rbind(ess,ESS)
  }
  return(list(xs=xs,ws=ws,ess=ess))
}

SVGPFfun<-function(data,N,m0,C0,alpha,beta,tau,r){
  if(missing(r)){r=2}else{}
  xs<-NULL
  ws<-NULL
  ess<-NULL
  x  = rnorm(N,m0,sqrt(C0))
  w  = rep(1/N,N)

    xprev<-x
    x<-rnorm(N,alpha+beta*x+(tau^2)/4*((data[t]^2)*exp(-(alpha+beta*x))-2),tau)
    w1<-w*dnorm(data[t],0,exp(x))*dnorm(x,alpha+beta*xprev,tau)/
      dnorm(x,mean=alpha+beta*xprev+(tau^2)/4*((data[t]^2)*exp(-(alpha+beta*xprev))-2),tau)

    xs = rbind(xs,x)
    ws = rbind(ws,w)
    ess =rbind(ess,ESS)
  }
  return(list(xs=xs,ws=ws,ess=ess))
}
#-------------------------------------------------------------------------

SVAPFoptfun<-function(data,N,m0,C0,alpha,beta,tau,r){
  if(missing(r)){r=2}else{}
  xs<-NULL
  ws<-NULL
  ess<-NULL
  x  = rnorm(N,m0,sqrt(C0))
  w  = rep(1/N,N)

    if(ESS<(N/r)){
      epsilon=alpha+beta*x+((tau^2)/4)*((data[t]^2)*exp(-alpha-beta*x)-2)
    weight = w*exp((1/(2*tau^2))*((epsilon)^2-(alpha+beta*x))-((data[t]^2)/2)*exp(-alpha-beta*x)*(1+alpha+beta*x))
    k   = sample(1:N,size=N,replace=TRUE,prob=weight)
    x1   = rnorm(N,alpha+beta*x[k]+(tau^2)/4*((data[t]^2)*exp(-(alpha+beta*x[k]))-2),tau)
    w   = rep(1/N,N)
    }else{
    weight = rep(1/N,N)
    k   = sample(1:N,size=N,replace=FALSE,prob=weight)
    x1   = rnorm(N,alpha+beta*x[k]+(tau^2)/4*((data[t]^2)*exp(-(alpha+beta*x[k]))-2),tau)
    w   <-  w*exp((1/(2*tau^2))*((alpha+beta*x[k]+((tau^2)/4)*((data[t]^2)*exp(-alpha-beta*x[k])-2))^2-(alpha+beta*x[k]))-((data[t]^2)/2)*exp(-alpha-beta*x[k])*(1+alpha+beta*x[k]))
    }

  }
  return(list(xs=xs,ws=ws,ess=ess))
}
#----------------------------------------------------------------------------------

SVLWfun<-function(data,N,m0,C0,ealpha,valpha,ebeta,vbeta,nu,lambda){

```{r include=FALSE}
N <- 5000
alpha <- -0.1
beta  <-  0.9
tau2  =  0.01
tau   = sqrt(tau2)
m0      = 0
C0      = 100
ealpha  = alpha
valpha  = 0.01
ebeta    = beta
vbeta    = 0.01
nu      = 3
lambda  = tau2

svpf<-SVPFfun(y,N,m0,C0, alpha,beta,tau)

svgpf<-SVGPFfun(y,N,m0,C0,alpha,beta,tau)

svapf<-SVAPFoptfun(y,N,m0,C0,alpha,beta=0,tau)

svlw<-SVLWfun(y,N,m0,C0,ealpha,valpha,ebeta,vbeta,nu,lambda)

SVFiltplot<-function(dataframe,fun,title,ub,lb){

ggplot(dataframe,aes(x=timeframe))+
  geom_line(aes(y=y), col="black", linetype=1)+
  geom_line(aes(y=x), col="black", linetype=3)+
  geom_line(aes(y=mean), col="red", linetype=1)+
  geom_ribbon(aes(ymin = mean-1.96*sd, ymax = mean+1.96*sd),
              fill="red",alpha=0.16) +
  #  scale_color_manual("",
   #                  values=c("Observations" = "black",
    #                         "True States" = "black",
     #                       "Filtered States"= "red"))+
  #scale_linetype_manual("",
   #                  values=c("Observations" = 1,
    #                         "True States" = 3,
     #                       "Filtered States"=1))+
  labs(x="Time",
       y="")+
  ggtitle(title)+
  coord_cartesian(ylim = c(lb, ub)) +
  theme_bw()+
#    theme(legend.direction = "horizontal", legend.position = "bottom", legend.key = element_blank(), 
#        legend.background = element_rect(fill = "white", colour = "gray30")) +
  theme(plot.title = element_text(hjust = 0.5))
}

We provide both a graphical and a numerical comparison and

```{r echo=FALSE, fig.align='center', fig.cap="Comparison of filtering strategies for Stochastic Volatility Model. Filtered States with credible interval (in red)", fig.height=6, fig.pos='H', fig.width=8, message=FALSE, warning=FALSE, results=F}
plotone<-SVFiltplot(dfsv,svpf,"",1.5,-2.5)
plottwo<-SVFiltplot(dfsv,svgpf,"",1.5,-2.5)
plotthree<-SVFiltplot(dfsv,svapf,"",1.5,-2.5)
plotfour<-SVFiltplot(dfsv,svlw,"",1.5,-2.5)

ggarrange(plotone, plottwo, plotthree, plotfour, nrow = 2, ncol = 2)
```

```{r}
set.seed(123)
truex<-x
estimatedx<-matrix(NA,ncol=5,nrow=4)
colnames(estimatedx)<-c("N","BPF","GPF", "APF","LWF")
estimatedx[,1]<-c(100,1000,10000,100000)

i=1
for(N in c(100,1000,10000,100000)){
set.seed(123)

svpf_1<-SVPFfun(y,N,m0,C0, alpha,beta,tau)
svgpf_1<-SVGPFfun(y,N,m0,C0,alpha,beta,tau)
svapf_1<-SVAPFoptfun(y,N,m0,C0,alpha,beta=0,tau)
svlw_1<-SVLWfun(y,N,m0,C0,ealpha,valpha,ebeta,vbeta,nu,lambda)

estimatedx[i,2]<-RMSE(truex,Filtervalues(svpf_1)$mean)
estimatedx[i,3]<-RMSE(truex,Filtervalues(svgpf_1)$mean)
estimatedx[i,4]<-RMSE(truex,Filtervalues(svapf_1)$mean)
estimatedx[i,5]<-RMSE(truex,Filtervalues(svlw_1)$mean)
i=i+1

@@

    xprev<-x
    x<-rnorm(N,alpha+beta*x+(tau^2)/4*((data[t]^2)*exp(-(alpha+beta*x))-2),tau)
    w1<-w*dnorm(data[t],0,exp(x))*dnorm(x,alpha+beta*xprev,tau)/
      dnorm(x,mean=alpha+beta*xprev+(tau^2)/4*((data[t]^2)*exp(-(alpha+beta*xprev))-2),tau)

@@

    if(ESS<(N/r)){
      epsilon=alpha+beta*x+((tau^2)/4)*((data[t]^2)*exp(-alpha-beta*x)-2)
    weight = w*exp((1/(2*tau^2))*((epsilon)^2-(alpha+beta*x))-((data[t]^2)/2)*exp(-alpha-beta*x)*(1+alpha+beta*x))
    k   = sample(1:N,size=N,replace=TRUE,prob=weight)
    x1   = rnorm(N,alpha+beta*x[k]+(tau^2)/4*((data[t]^2)*exp(-(alpha+beta*x[k]))-2),tau)
    w   = rep(1/N,N)
    }else{
    weight = rep(1/N,N)
    k   = sample(1:N,size=N,replace=FALSE,prob=weight)
    x1   = rnorm(N,alpha+beta*x[k]+(tau^2)/4*((data[t]^2)*exp(-(alpha+beta*x[k]))-2),tau)
    w   <-  w*exp((1/(2*tau^2))*((alpha+beta*x[k]+((tau^2)/4)*((data[t]^2)*exp(-alpha-beta*x[k])-2))^2-(alpha+beta*x[k]))-((data[t]^2)/2)*exp(-alpha-beta*x[k])*(1+alpha+beta*x[k]))
    }

@@

  }
  return(list(xs=xs,ws=ws,ess=ess))
}
